Making Sequential Programs Deterministic

### 2019-04-05 design.tex
Edits in section 3.1
types for get and set
Polymorphism syntax in Java

@@

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine, as shown in 
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible element type for a \<Det> collection is \<Det>.

@@

\subsection{Determinism checker specific rules for precision}\label{precision}

@@

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

>>>>

    \rulename{get}: 
    $
    \forallt{\tau} \kappa \in \{\<Det>, \<NonDet>\}\ \<OrderNonDet> \ C \angles{\tau} \ \times \kappa \ \<Int> \rightarrow 
    \<NonDet> \ \<Int>
    $

    \rulename{set}: 
    $
    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<Det> \ \<Int> \times \kappa \ \beta_n \rightarrow 
    \<NonDet> \ \beta
    $

    \rulename{set invalid}: 
    $
    \forallt{\kappa, \beta} \ \<OrderNonDet> \ C \angles{\kappa \ \beta} \ \times \<NonDet> \ \<Int> \times \kappa \ \beta_n \rightarrow 
    \|INVALID|
    $

@@

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine,
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible element type for a \<Det> collection is \<Det>.

@@

\subsection{Improving precision}\label{precision}

@@

We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("up") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y)| : \<Det> \ \|boolean|}{\Gamma \vdash \|x| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \<OrderNonDet> \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \<OrderNonDet> \ \<List> & \kappa_{y} \ \beta_{y} \nsqsubseteq \<OrderNonDet> \ \<List> }$

### 2019-04-05 java-design.tex
Polymorphism syntax in Java

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

\todo{Explain @PolyDet("up"), "down", "use".}

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could be instantiated
%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det} or \codeid{@NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> gets instantiated to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

>>>>

\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity> is $\forallt{\tau} \tau \rightarrow \tau$, and
the type of \<identity(arg)> is the same as the type of
\<arg>.  A programmer can also write a type qualifier on a type argument,
as in \codeid{@NonDet List<@NonDet Integer>}.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
interface ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so \theDeterminismChecker uses a special type qualifier name, \<@PolyDet>.  (\<@PolyDet> stands
for ``polymorphic determinism qualifier.)
A qualifier-polymorphic method \<m> with signature $\forallt{\kappa} \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

Qualifier  polymorphism is common on methods that a programmer might think
of as deterministic.  For example, an addition method could be defined as

\begin{verbatim}
    @PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}

This can be used in more contexts than

\begin{verbatim}
    @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{verbatim}

\noindent
can be, as was shown in \cref{sec:basic-polymorphism}.

Just as a qualifier variable $\kappa$ is written as \<@PolyDet> in Java
source code, $\up{\kappa}$ is written as \<@PolyDet("up")>, and
$\down{\kappa}$ is written as \<@PolyDet("down")>.  An occurrence of a
qualifier variable that does not affect the binding of that variable
(\cref{bindings-uses}) is written \<@PolyDet("use")>.
